---
title: "Homicides Outline"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(WDI)
library(wbstats)
library(tidyverse)
library(ggalt)
```

```{r}
df_region <- read_csv("Data/un-regional-homicide-rate.csv")
df_region %>% 
  filter(year == 2018) %>% 
  select(un_region, rate) %>% 
  ggplot(aes(x= un_region, y = rate, fill = un_region)) +
  geom_bar(stat="identity")
```

Americas has the most by far with Africa as a close-ish runner up. Other regions fall way below (I wonder if Americas is lowered by North America- how is it weighted?)


```{r}
df_income <- wbstats::wb_data(
  # Incomelevels
  country = wb_income_levels()$iso3c,
  indicator = c('VC.IHR.PSRC.P5', 'VC.IHR.PSRC.FE.P5', 'VC.IHR.PSRC.MA.P5'),
  return_wide = TRUE)
df_income %>% 
  filter(date == 2015) %>% 
  select(iso3c, VC.IHR.PSRC.P5, country) %>% 
  ggplot(aes(x= iso3c, y = VC.IHR.PSRC.P5, fill = country)) +
  geom_bar(stat="identity")
```


```{r}
# pull and combine data
intentional_homicide = WDI(indicator='VC.IHR.PSRC.P5', extra = TRUE)
intentional_homicide_fem = WDI(indicator = "VC.IHR.PSRC.FE.P5", extra = TRUE)
intentional_homicide_male = WDI(indicator = "VC.IHR.PSRC.MA.P5", extra = TRUE)

intentional_homicide["Female"] <- intentional_homicide_fem["VC.IHR.PSRC.FE.P5"]
intentional_homicide["Male"] <- intentional_homicide_male["VC.IHR.PSRC.MA.P5"]
```

```{r}
intentional_homicide_map <- WDI(indicator='VC.IHR.PSRC.P5', extra = TRUE, latest = 1) %>% 
  write_csv("Data/homicides_map.csv")
```

https://public.flourish.studio/visualisation/6058072/


```{r}
intentional_homicide_MRV = WDI(indicator='VC.IHR.PSRC.P5', extra = TRUE, latest = 1)
intentional_homicide_fem_MRV = WDI(indicator = "VC.IHR.PSRC.FE.P5", extra = TRUE, latest = 1)
intentional_homicide_male_MRV = WDI(indicator = "VC.IHR.PSRC.MA.P5", extra = TRUE, latest = 1)

intentional_homicide_gendered_MRV <- merge(intentional_homicide_fem_MRV, intentional_homicide_male_MRV, by = "country") %>% 
  rename(Female = VC.IHR.PSRC.FE.P5,
         Male = VC.IHR.PSRC.MA.P5)
```


```{r}
# but when we look at the world aggregates
df_region %>% 
  select(un_region, rate, year) %>% 
  ggplot(aes(x= year, y = rate, color = un_region)) +
  geom_point(stat="identity")
```

- There's a clear downward trend since data collection begins in 1990. 
- However these gains have not been distributed evenly, and some regions continue to have much higher homicide rates. Additionally there is a very noticeable gender discrepancy
- Americas is the only one that seems to stay up

```{r}
# put them both on the same graph
intentional_homicide %>% 
  pivot_longer(cols = c(Male, Female)) %>% 
  drop_na() %>% 
  ggplot(aes(x= year, y = value, color=name)) +
  geom_smooth(method = "loess",se=F,size=1.5) +
  facet_wrap(~name,nrow=1) +
  geom_point() 

mean_male <- mean(intentional_homicide$Male, na.rm = TRUE)

mean_female <- mean(intentional_homicide$Female, na.rm = TRUE)


# density plot
intentional_homicide %>% 
  pivot_longer(cols = c(Male, Female)) %>% 
  drop_na() %>% 
  ggplot(aes(x=value, fill = name)) +
  geom_density(alpha=0.4) +
    geom_vline(aes(xintercept=mean_male), color = "blue",
             linetype="dashed") +
  geom_vline(aes(xintercept=mean_female), color = "red",
             linetype="dashed") +
  xlim(0,50)
```

As demonstrated by the charts above, men and women have drastically different homicide rates, both in terms of averages and distributions. 

Women are much more highly concentrated at lower homicide rates
Men are much more evenly distributed and the distirbution extends all the way to 200
The mean of men is also signficantly higher (14.2 as compared to 2.3)


```{r}
# by region
intentional_homicide %>%
  filter(region != 'Aggregates',year > 1990) %>% 
  ggplot(aes(x= year, y = VC.IHR.PSRC.P5, color = region)) +
  geom_point()

# averages
intentional_homicide %>% 
  filter(region != "Aggregates", year > 1990) %>% 
  drop_na() %>% 
  group_by(year, region) %>% 
  summarise(VC.IHR.PSRC.P5 = mean(VC.IHR.PSRC.P5)) %>% 
  ggplot(aes(x= year, y = VC.IHR.PSRC.P5, color = region)) +
  geom_point()

# averages
intentional_homicide %>% 
  filter(region == "North America", year > 1990) %>% 
  drop_na() %>% 
  group_by(year, region) %>% 
  summarise(VC.IHR.PSRC.P5 = mean(VC.IHR.PSRC.P5)) %>% 
  ggplot(aes(x= year, y = VC.IHR.PSRC.P5, color = region)) +
  geom_point()
```
The first graph shows individual countries colored by their respective region. 
The second graph shows regional averages
Latin America has significantly higher averages than any other region
Sub Saharan Africa spikes after 2005?
North America also doens't have the same pronounced decrease

```{r}
# bar chart by region in 2000
intentional_homicide %>% 
  filter(region != "Aggregates", year == 2000) %>% 
  drop_na() %>% 
  group_by(region) %>% 
  summarise(VC.IHR.PSRC.P5 = mean(VC.IHR.PSRC.P5)) %>% 
  ggplot(aes(x= region, y = VC.IHR.PSRC.P5, fill = region)) +
  geom_bar(stat="identity")

# bar chart by region in 2010
intentional_homicide %>% 
  filter(region != "Aggregates", year == 2010) %>% 
  drop_na() %>% 
  group_by(region) %>% 
  summarise(VC.IHR.PSRC.P5 = mean(VC.IHR.PSRC.P5)) %>% 
  ggplot(aes(x= region, y = VC.IHR.PSRC.P5, fill = region)) +
  geom_bar(stat="identity")

#bar chart by region in 2018
intentional_homicide %>% 
  filter(region != "Aggregates", year == 2018) %>% 
  drop_na() %>% 
  group_by(region) %>% 
  summarise(VC.IHR.PSRC.P5 = mean(VC.IHR.PSRC.P5)) %>% 
  ggplot(aes(x= region, y = VC.IHR.PSRC.P5, fill = region)) +
  geom_bar(stat="identity")
```
Again, Latin America is noticeably higher
North America also doesn't seem to experience the same gains
Animated or interactive that goes year by year and shows the change over time


```{r}
# break up by income group
intentional_homicide %>%
  filter(income != 'Aggregates',year > 1990) %>%
  ggplot(aes(x= year, y = VC.IHR.PSRC.P5, color = income)) +
  geom_point()
```
. 
Interestingly it seems like the relationship between income and homicides are different than we may expect. Middle incomes seem to have the highest rates, high income is consistently low, and low income varies.

Middle incomes have higher medians. Upper middle income actually has the highest median but lower middle income has a greater distirbution

Finding: High income reduces homicide rates as expected but middle income seems to really be where the conflict is. 


```{r}
# regress homicides on income
intentional_homicide %>% 
  filter(region != "Aggregates", year > 1990) %>% 
  drop_na() %>% 
  lm(VC.IHR.PSRC.P5 ~ factor(income), data = .) %>% 
  summary()
```

Based on the regression it seems like upper middle income and lower middle income are actually more likely to be correlated with homicides and are highly statistically significant. 

```{r, fig.height= 20, fig.width= 10}
# dumbbell plot

intentional_homicide_gendered_MRV %>% 
  mutate(country = fct_reorder(country, Male)) %>% 
  filter(region.x != "Aggregates") %>% 
  ggplot(aes(x=Female, xend=Male, y=country)) + 
    geom_dumbbell(size=1.5, color="#b2b2b2", size_x=3, size_xend = 3, colour_x = 'red', colour_xend = 'blue') +
    geom_text(aes(x=Male, y=country, label="Male"),
          color='blue', size=3, vjust=-1.5) +
  geom_text(aes(x=Female, y=country, label="Female"),
          color='red', size=3, vjust=-1.5) +
  geom_text(aes(x=Male, y=country, label=round(Male, 1)),
          color='blue', size=2.75, vjust=2.5) +
  geom_text(color='red', size=2.75, vjust=2.5,
          aes(x=Female, y=country, label=round(Female, 1))) 
  
    
```

Even in Latin America being a woman reduces your homicide possibilty significanlty. 
Middle East low in both. Man in East Asia better than woman in LAC. 
The two seem to track together -> high homicide rate in one generally means high homicide rate in the other. Non indepdnece means intervention will need to address undelrying factors causing both.  



