---
title: "Conflict and Homicides"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages


```{r}
library(WDI)
library(wbstats)
library(tidyverse)
```

## Homicides 


```{r}
# pull and combine data
intentional_homicide = WDI(indicator='VC.IHR.PSRC.P5')
intentional_homicide_fem = WDI(indicator = "VC.IHR.PSRC.FE.P5")
intentional_homicide_male = WDI(indicator = "VC.IHR.PSRC.MA.P5")

intentional_homicide["Female"] <- intentional_homicide_fem["VC.IHR.PSRC.FE.P5"]
intentional_homicide["Male"] <- intentional_homicide_male["VC.IHR.PSRC.MA.P5"]
```

```{r}
intentional_homicide <- intentional_homicide %>% 
  drop_na()
```


```{r}
intentional_homicide %>% 
  ggplot(aes(x= year, y = VC.IHR.PSRC.P5)) +
  geom_point() 
```

```{r}
intentional_homicide %>% 
  slice(-(1:2868)) %>% 
  group_by(year) %>% 
  drop_na() %>% 
  summarise(VC.IHR.PSRC.P5 = sum(VC.IHR.PSRC.P5)) %>% 
  ggplot(aes(x= year, y = VC.IHR.PSRC.P5)) +
  geom_point()

intentional_homicide %>% 
  group_by(year) %>% 
  drop_na() %>% 
  filter(country == "World") %>% 
  ggplot(aes(x= year, y = VC.IHR.PSRC.P5)) +
  geom_point()
```

- Homicides go up (potentially data collection) and then come back down

_ But when we do world the trend is downward? Is there somehting about the data collection that I'm not understanding

```{r}
intentional_homicide %>% 
  group_by(year) %>% 
  drop_na() %>% 
  summarise(Female = sum(Female)) %>% 
  ggplot(aes(x= year, y = Female)) +
  geom_point() 

intentional_homicide %>% 
  group_by(year) %>% 
  drop_na() %>% 
  summarise(Male = sum(Male)) %>% 
  ggplot(aes(x= year, y = Male)) +
  geom_point()


```

- men commit way more homicides than women

```{r}
# put them both on the same graph
intentional_homicide %>% 
  pivot_longer(cols = c(Male, Female)) %>% 
  drop_na() %>% 
  ggplot(aes(x= year, y = value, color=name)) +
  geom_smooth(method = "loess",se=F,size=1.5) +
  facet_wrap(~name,nrow=1) +
  geom_point() 
```

```{r}
# by region
intentional_homicide %>%
  filter(country %in% c("Arab World", "East Asia & Pacific", "Europe & Central Asia", "Latin America & Caribbean"), year > 2000) %>% 
  ggplot(aes(x= year, y = VC.IHR.PSRC.P5, color = country)) +
  geom_point()
  
```
```{r}
# break up by income group
income <- c("Low income", "Middle income", "High income")

intentional_homicide %>%
  filter(country %in% income) %>% 
  ggplot(aes(x= year, y = VC.IHR.PSRC.P5, color = country)) +
  geom_point()
 
```

-LAC has super high homicide rates
- High income has lower homicides than medium, low


## Conflict

```{r}
# WDI Battle Deaths
battle_deaths <- WDI(indicator = 'VC.BTL.DETH') 

# dropna
battle_deaths<- battle_deaths %>% 
  drop_na()
```



```{r}
battle_deaths %>% 
  slice(-(1:183)) %>% 
  ggplot(aes(x= year, y = VC.BTL.DETH)) +
  geom_point() +
  geom_text(aes(label=ifelse(VC.BTL.DETH>75000,as.character(country),'')),hjust=0,vjust=0, size = 3)

battle_deaths %>% 
  slice(-(1:183)) %>% 
  group_by(year) %>% 
  drop_na() %>% 
  summarise(VC.BTL.DETH = sum(VC.BTL.DETH)) %>% 
  ggplot(aes(x= year, y = VC.BTL.DETH)) +
  geom_smooth(method = "loess",se=F,color="red",size=.5) +
  geom_point() 
```
- There's a clear trend up that goes down, back up, and then peaks at about 2015 and then comes down again

```{r}
# remove syria
battle_deaths %>% 
  slice(-(1:183)) %>% 
  filter(country != "Syrian Arab Republic") %>% 
  group_by(year) %>% 
  drop_na() %>% 
  summarise(VC.BTL.DETH = sum(VC.BTL.DETH)) %>% 
  ggplot(aes(x= year, y = VC.BTL.DETH)) +
  geom_smooth(method = "loess",se=F,color="red",size=.5) +
  geom_point() 


```
- filtering by Syria gets rid of the peak, but there is still a pronounced increase from 2010 on

```{r warning = FALSE, message = FALSE}
#ACLED
setwd("Data")
ACLED <- read_csv("ACLED.csv")
  
```

```{r}
ACLED %>% 
  count(event_type)
```

```{r}
ACLED %>% 
  ggplot(aes(event_type)) +
  geom_bar()

ACLED %>% 
  ggplot(aes(event_type, fatalities, fill = region)) +
  geom_col()

ACLED %>% 
  ggplot(aes(event_type, fatalities, fill = region)) +
  geom_col(position = "dodge")
```

- Protests are by far the most common occurence, with battles second. Battles are the most deadly by far




```{r}
ACLED %>% 
  ggplot(aes(region, fatalities, fill = region)) +
  geom_col()
```

Middle East has the largest number of fatalities in the ACLED data

